---
title: "Evagene risk explorer (R / Quarto)"
author: "Evagene integration examples"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    embed-resources: true
execute:
  echo: true
  warning: false
---

A **what-if** tour of Evagene's risk outputs — one code chunk per question,
one chart per answer. The narrative mirrors the Python / Jupyter variant;
pick whichever kernel matches your toolchain.

This notebook is a Quarto document. Render it with `quarto render
explorer.qmd` (produces `explorer.html`) or preview interactively with
`quarto preview explorer.qmd`. Full widget-style interactivity in a
rendered HTML file requires [Shinylive](https://quarto-ext.github.io/shinylive/)
or `quarto preview`; the static render below re-executes with a modest
parameter grid so the charts still read cleanly.

**The scratch pedigree is created in Section 4 and deleted in Section 7.**
If the document aborts mid-way, find the orphan in the Evagene web UI
(it is named `[scratch] notebook-explorer ...`) and delete it by hand.

## 1. Setup

The notebook uses a tiny R module (`R/client.R`) whose only job is to
make the chunks below short. The key lives in an environment variable —
never in a chunk.

> **First-time setup:** copy `.Renviron.example` to `.Renviron` in this
> folder, paste in your Evagene API key (scopes `read` + `write` +
> `analyze`), and restart R.

```{r setup, message=FALSE}
r_dir <- normalizePath("R", mustWork = TRUE)
for (f in list.files(r_dir, pattern = "\\.R$", full.names = TRUE)) {
  source(f, local = FALSE)
}

# Walk up a few levels looking for a .env file.  Convenient when the
# notebook is rendered from a fresh R session.
load_dotenv_walking <- function() {
  here <- normalizePath(".", mustWork = TRUE)
  for (i in seq_len(5)) {
    candidate <- file.path(here, ".env")
    if (file.exists(candidate)) {
      lines <- readLines(candidate, warn = FALSE)
      for (line in lines) {
        line <- trimws(line)
        if (!nzchar(line) || startsWith(line, "#") || !grepl("=", line)) next
        parts <- strsplit(line, "=", fixed = TRUE)[[1L]]
        key <- trimws(parts[[1L]])
        value <- trimws(paste(parts[-1L], collapse = "="))
        if (Sys.getenv(key, unset = "") == "") {
          do.call(Sys.setenv, stats::setNames(list(value), key))
        }
      }
      return(invisible(NULL))
    }
    parent <- dirname(here)
    if (parent == here) break
    here <- parent
  }
}
load_dotenv_walking()

library(ggplot2)
library(knitr)

config <- load_config(as.list(Sys.getenv()))
gateway <- make_httr2_gateway()
client <- evagene_client(gateway, config$base_url, config$api_key)
sprintf("Ready. Base URL: %s", config$base_url)
```

## 2. Pick a pedigree

The list is pulled from `GET /api/pedigrees`. Each row links back to the
web UI.

```{r list-pedigrees, results='asis'}
pedigrees <- client$get_pedigrees()
rows <- lapply(pedigrees, function(p) {
  data.frame(
    Name = p$display_name %||% "(untitled)",
    Link = sprintf("[open](%s)", client$evagene_url(p$id)),
    ID = sprintf("`%s`", p$id),
    stringsAsFactors = FALSE
  )
})
knitr::kable(do.call(rbind, rows))
```

Default: prefer a pedigree whose display name mentions "breast" (so the
NICE and Tyrer-Cuzick sections produce meaningful numbers); fall back to
the first pedigree otherwise. Override by setting `PEDIGREE_ID` below.

```{r pick-pedigree}
selected_idx <- which(
  vapply(pedigrees, function(p) grepl("breast", tolower(p$display_name %||% "")), logical(1L))
)[1]
if (is.na(selected_idx)) selected_idx <- 1L
selected <- pedigrees[[selected_idx]]
PEDIGREE_ID <- selected$id
cat(sprintf("Selected: %s\n", selected$display_name))
cat(sprintf("Link:     %s\n", client$evagene_url(PEDIGREE_ID)))
```

## 3. Baseline NICE category

One call to `POST /api/pedigrees/{id}/risk/calculate` with `model=NICE`.
The response carries `cancer_risk$nice_category` (one of `near_population`,
`moderate`, `high`), the triggers that fired, and human-readable notes.

```{r baseline-nice}
TRAFFIC_LIGHTS <- c(
  near_population = "GREEN",
  moderate = "AMBER",
  high = "RED"
)

summarise_nice <- function(payload) {
  cr <- payload$cancer_risk %||% list()
  category <- cr$nice_category %||% "unknown"
  triggers <- cr$nice_triggers %||% list()
  notes <- cr$notes %||% list()
  light <- TRAFFIC_LIGHTS[category] %||% "unknown"
  paste(
    sprintf("NICE category: %s (%s)", toupper(category), light),
    sprintf("Triggers: %s",
            if (length(triggers) == 0L) "none" else paste(unlist(triggers), collapse = "; ")),
    if (length(notes) > 0L) sprintf("Note: %s", notes[[1L]]) else NULL,
    sep = "\n"
  )
}

baseline_nice <- client$run_risk(PEDIGREE_ID, "NICE")
cat(summarise_nice(baseline_nice))
```

## 4. What if we add affected first-degree relatives?

This is where the scratch clone comes in. The notebook:

1. Clones the pedigree via a GEDCOM round-trip (export → create → import).
2. Adds a synthetic sister to the proband; flags the sister and (if
   present) the mother affected with breast cancer.
3. Re-runs the NICE calculation on the scratch copy.

Two affected first-degree relatives is the textbook NICE "moderate"
(amber) trigger.

```{r clone-pedigree}
scratch_suffix <- format(Sys.time(), "%Y-%m-%d %H:%M UTC", tz = "UTC")
SCRATCH_ID <- client$clone_pedigree_for_exploration(PEDIGREE_ID, scratch_suffix)
cat(sprintf("Scratch pedigree: %s\n", client$evagene_url(SCRATCH_ID)))
```

```{r pick-disease}
# We need the scratch proband's ID and a "Breast Cancer" disease ID.
register <- client$get_register(SCRATCH_ID)
proband_id <- register$proband_id
cat(sprintf("Scratch proband: %s\n", proband_id))

# Look up the breast-cancer disease by display name in the catalogue,
# so the demo works regardless of which diseases were attached to the
# pedigree you picked.
diseases_request <- list(
  method = "GET",
  url = sprintf("%s/api/diseases", config$base_url),
  headers = c("X-API-Key" = config$api_key, "Accept" = "application/json")
)
disease_catalogue <- jsonlite::fromJSON(
  gateway(diseases_request)$body_string, simplifyVector = FALSE
)
breast_cancer <- NULL
for (d in disease_catalogue) {
  if (identical(trimws(d$display_name %||% ""), "Breast Cancer")) {
    breast_cancer <- d
    break
  }
}
if (is.null(breast_cancer)) {
  stop("No 'Breast Cancer' disease in your catalogue; cannot continue section 4.")
}
BREAST_CANCER_ID <- breast_cancer$id
cat(sprintf("Breast Cancer disease id: %s\n", BREAST_CANCER_ID))
client$add_disease_to_pedigree(SCRATCH_ID, BREAST_CANCER_ID)
```

```{r add-affected}
sister <- client$add_relative(
  SCRATCH_ID,
  relative_of = proband_id,
  relative_type = "sister",
  display_name = "Synthetic sister",
  biological_sex = "female"
)
sister_id <- sister$individual$id
cat(sprintf("Sister id: %s\n", sister_id))

client$add_disease_to_individual(
  sister_id,
  disease_id = BREAST_CANCER_ID,
  age_at_diagnosis = 38L
)

# Also mark the mother affected (two affected first-degree relatives is
# the textbook NICE "moderate" trigger — one affected sister alone does
# not reliably cross the threshold for every family structure).
mother_row <- NULL
for (row in register$rows) {
  if (identical(row$relationship_to_proband, "Mother")) {
    mother_row <- row
    break
  }
}
if (!is.null(mother_row)) {
  client$add_disease_to_individual(
    mother_row$individual_id,
    disease_id = BREAST_CANCER_ID,
    age_at_diagnosis = 45L
  )
  cat("Marked sister and mother affected with Breast Cancer.\n")
} else {
  cat("Marked sister affected with Breast Cancer (no mother in the pedigree).\n")
}
```

```{r rerun-nice}
after_nice <- client$run_risk(SCRATCH_ID, "NICE")
cat("BEFORE:\n")
cat(summarise_nice(baseline_nice), "\n\n")
cat("AFTER:\n")
cat(summarise_nice(after_nice))
```

## 5. Tyrer-Cuzick slider

`TYRER_CUZICK` reads the **proband's** reproductive-history fields:
`age_at_menarche`, `parity`, `age_at_first_live_birth`,
`breast_density_birads` (BI-RADS 1–4), `hormone_therapy_years`, and so
on. They live on the individual record, not the risk-request body — so
this chunk mutates them on the scratch proband and re-runs TC for each
combination. The grid is deliberately small (so the chunk executes
inside a minute); for richer sweeps, increase the vectors below when
rendering locally. The `httr2` gateway retries on HTTP 429 transparently.

```{r tc-sweep}
MENARCHE_AGES <- c(10L, 13L)
PARITIES <- c(0L, 2L)
BIRADS_LEVELS <- 1:4

tc_results <- list()
for (menarche in MENARCHE_AGES) {
  for (parity in PARITIES) {
    for (birads in BIRADS_LEVELS) {
      client$patch_individual(
        proband_id,
        fields = list(
          age_at_menarche = menarche,
          parity = parity,
          breast_density_birads = birads,
          age_at_first_live_birth = if (parity > 0) 28L else NULL
        )
      )
      Sys.sleep(0.4)  # headroom under the 60/min limit
      tc <- client$run_risk(SCRATCH_ID, "TYRER_CUZICK")
      cancer_risk <- tc$cancer_risk %||% list()
      tc_results[[length(tc_results) + 1L]] <- data.frame(
        menarche = menarche,
        parity = parity,
        birads = birads,
        ten_year = cancer_risk$tc_ten_year_risk %||% NA_real_,
        lifetime = cancer_risk$tc_lifetime_risk %||% NA_real_,
        stringsAsFactors = FALSE
      )
    }
  }
}
tc_df <- do.call(rbind, tc_results)
tc_df$series <- paste0("menarche ", tc_df$menarche, ", parity ", tc_df$parity)
sprintf("TC results: %d points", nrow(tc_df))
```

```{r tc-plot, fig.width=8, fig.height=5}
ggplot(tc_df, aes(x = birads, y = ten_year * 100, colour = series)) +
  geom_line() + geom_point(size = 2) +
  scale_x_continuous(breaks = 1:4) +
  labs(
    x = "BI-RADS breast density (1 = almost entirely fatty, 4 = extremely dense)",
    y = "Tyrer-Cuzick 10-year risk (%)",
    colour = NULL,
    title = "Tyrer-Cuzick 10-year risk vs breast density, by reproductive history"
  ) +
  theme_minimal()
```

Reading the chart: each curve holds `age_at_menarche` and `parity`
constant while BI-RADS walks from 1 (fatty) to 4 (extremely dense).
Denser breasts drag risk up steeply; higher parity (with age at first
live birth fixed at 28) shifts the whole curve down — the classic
protective effect.

## 6. Multifactorial under varied heritability

The multifactorial (Falconer liability-threshold) model takes a disease's
`heritability` ($h^2$) and `population_prevalence` ($K$), plus the
affected-relative structure, and returns a recurrence risk for the
proband.

`/risk/calculate` does **not** accept a request-level $h^2$ override, so
to trace the curve the notebook runs the model once per disease against
the (sister-is-affected) scratch family. Each disease is a point on the
$h^2$ axis.

```{r pick-heritable}
heritable <- Filter(function(d) !is.null(d$heritability), disease_catalogue)
by_name <- setNames(heritable, vapply(heritable, `[[`, character(1L), "display_name"))
NAMES <- c(
  "Major Depressive Disorder (Familial)",
  "Type 2 Diabetes (Familial)",
  "Epilepsy (Familial)",
  "Cleft Lip With or Without Cleft Palate",
  "Schizophrenia (Familial)",
  "Bipolar Disorder (Familial)",
  "Coeliac Disease",
  "Ankylosing Spondylitis"
)
picks <- by_name[intersect(NAMES, names(by_name))]
sprintf("%d heritable diseases available; using %d for the plot.",
        length(heritable), length(picks))
```

```{r mf-sweep}
mf_rows <- list()
for (d in picks) {
  did <- d$id
  client$add_disease_to_pedigree(SCRATCH_ID, did)
  Sys.sleep(0.3)
  client$add_disease_to_individual(
    sister_id, disease_id = did, age_at_diagnosis = 30L
  )
  Sys.sleep(0.3)
  out <- client$run_risk(SCRATCH_ID, "MULTIFACTORIAL", extra = list(disease_id = did))
  block <- out$multifactorial %||% list()
  mf_rows[[length(mf_rows) + 1L]] <- data.frame(
    disease = d$display_name,
    h2 = d$heritability,
    K = d$population_prevalence %||% NA_real_,
    final_risk = block$final_risk %||% NA_real_,
    base_risk = block$base_risk %||% NA_real_,
    stringsAsFactors = FALSE
  )
}
mf_df <- do.call(rbind, mf_rows)
knitr::kable(mf_df)
```

```{r mf-plot, fig.width=8, fig.height=5}
ggplot(mf_df, aes(x = h2, y = final_risk * 100)) +
  geom_point(size = 3) +
  geom_text(aes(label = disease), nudge_y = 1.5, size = 3, check_overlap = TRUE) +
  labs(
    x = "Disease heritability of liability (h²)",
    y = "Multifactorial recurrence risk for proband (%)\nwith one affected first-degree relative",
    title = paste(
      "Recurrence risk rises with heritability under",
      "the Falconer liability-threshold model"
    )
  ) +
  theme_minimal()
```

Each point is a real disease, run through Evagene's multifactorial
engine against the same synthetic family structure (one affected
first-degree relative, otherwise unaffected). Population prevalence
varies across diseases and also drives the curve — the Falconer model
is a joint function of $h^2$ and $K$, so the points do not fall on a
single line. That is the teaching point: heritability is a lever, not
a knob.

## 7. Caveats and cleanup

The final chunk deletes the scratch pedigree. If a preceding chunk
errored, knit will stop before reaching here — run this chunk manually
before retrying, otherwise the scratch copy stays in your account.

```{r cleanup}
client$delete_pedigree(SCRATCH_ID)
cat(sprintf(
  "Scratch pedigree %s deleted. Your original is untouched.\n", SCRATCH_ID
))
```

### Caveats

- **Tyrer-Cuzick** is an IBIS-style approximation, not the licensed
  IBIS tool. **BOADICEA** is not bundled; for the full multi-gene /
  polygenic-risk-score version, export a CanRisk v2 file from the
  pedigree and upload at [canrisk.org](https://canrisk.org).
- **Multifactorial recurrence risk** shown here is Evagene's Falconer
  liability-threshold output with Carter-effect and consanguinity
  modifiers. Published empirical recurrence tables (when they exist)
  are a better source for the specific disease.
- These are example integrations, **not validated clinical tools**.

### Suggested reading

- Falconer DS. *Ann Hum Genet* 1965 — the liability-threshold foundation.
- Tyrer J, Duffy SW, Cuzick J. *Stat Med* 2004 — the IBIS breast-cancer
  model.
- NICE CG164 / NG101 — family-history triage rules driving the NICE
  category shown in Section 3.
- Parmigiani et al. *Am J Hum Genet* 1998 — BRCAPRO, the Bayesian
  counterpart.
